Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
action on my TODO list

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
%

@@

\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  We also investigate more complex supervised
approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
similarly complex phenotype that requires higher order dependencies between features to be taken into
account. %, as done by RF and DL.
\marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
  to the multi variable interactions and RF benefit of being hard to overfit).  Using conventional machine learning
  analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model
  over such a high-dimensional feature space~\cite{Chen2012}.  As a result machine learning methods such as random
  forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make
  the approach robust against overfitting issues~\cite{Breiman2001}.  However, training a supervised learning method on
  such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel
  approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models
  based on information from the full dataset.}

@@

\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper)}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
Scotland).

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate
classification into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.936 \pm{} 0.014\).
This is an improvement on our previous result from \kMeans{} clustering the same dataset in our previous work, where we observed an ARI of \(0.83\).

To determine where the model is underperforming, we also record the precision and recall for each of the groups.
We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
Consequently, this led to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
Although these misclassifications as AMR are relatively frequent, we didn't observe any AFR individuals being classified as EUR, or vice-versa, in any of the ten folds.
Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
These results demonstrate the diversity, or relative difference of some of the groups.

>>>>

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
\usepackage{placeins}
%

@@

\marginpar{TODO do we need to cite ~\cite{Bureau2005,Yoo2012,Qi2012}.}  
We also investigate more complex supervised
approaches, namely Random Forest (RF), as for example Locke {\it et al.}~\cite{Locke2015} showed in a recent meta-study
around genetic susceptibility of obesity that only five of the 97 genomic loci have an independent association signal
the rest seem to jointly predict the complex metabolic phenotype.  We therefore investigate whether ethnicity is a
similarly complex phenotype that requires higher order dependencies between features to be taken into
account. %, as done by RF and DL.

RF inherit many of the benefits of decision trees. It handles large numbers of variables transparently and
provides an unbiased estimate of the generalization error (in addition, it handles missing values and unbalanced
population classes sizes). They provide a measure of variable importance and allow for interactions between
variables to be investigated. However, in this problem, with many millions of variables, inspecting plots of interactions
is not feasible.  We note that decision trees (and random forests) model local variable interactions -- there are 
no main effects or global interaction effects.

For shallow neural networks there are a number of methods for controlling overfitting including regularization, max norm
constraints, and dropout. Whether the problem of overfitting is exacerbated or not in deep networks is an open problem.
With a large number of variables (like this problem) some regularization is going to be necesssary. 
%http://stats.stackexchange.com/questions/21152/obtaining-knowledge-from-a-random-forest

\marginpar{TODO Rob do you want to add two sentences about RF and DL and the respective benefits (ideally with respect
  to the multi variable interactions and RF benefit of being hard to overfit).  Using conventional machine learning
  analysis to elucidate this additive effects is challenging as there are not enough samples to robustly fit the model
  over such a high-dimensional feature space~\cite{Chen2012}.  As a result machine learning methods such as random
  forests (RF) have become popular as the underlying decision trees are build from a subsets of the features, which make
  the approach robust against overfitting issues~\cite{Breiman2001}.  However, training a supervised learning method on
  such an expansive dataset as the 1000 genomes is computationally challenging and traditional massively parallel
  approaches to data processing are not easily applicable to machine learning tasks, as they iteratively refine models
  based on information from the full dataset.}

@@

\FloatBarrier
\clearpage
\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper)}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
Scotland).

@@

\subsubsection{Logistic regression is more accurate than \kMeans{} in capturing the full population structure}
In this section we train and test our two methods on the full dataset of 14 populations and assess the accuracy with respect to the accurate classification into the 4 super-populations.
Using 10-fold cross-validation to assess the logistic regression classification results, we observe an average ARI of \(0.93 \pm{} XX\). This is an improvement on our previous result from \kMeans{} clustering the same dataset
where we observed an ARI of \(0.83\).
We also record the precision and recall for each of the groups, and observe that for each classification, and both values are always 1.0 for the EAS super-population. This demonstrates that each logistic regression model was able to classify every EAS individual correctly while not including any false positives in this group.
For the AMR group, however, we consistently observed a relatively low precision (\(0.85\pm{} XX \)). This was generally due to individuals from the EUR super-population being misclassified as AMR. As a result, this led to EUR having a relatively low recall (\(0.9 \pm XX \)).
Although not present in the result of every fold, another observation is that AFR individuals would sometimes be misclassified as AMR. In all ten folds, we didn't observe AFR being classified as EUR, and vice-verse, nor AMR individuals being misclassified as another super-population.
These results demonstrate the diversity, or relative difference of some of the groups.
